---
title: "Multicollinearity: Question"
subtitle: "STA 506 2.0 Linear Regression Analysis"
author: "Thiyanga S. Talagala"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

```{r, comment=NA, message=FALSE, warning=FALSE}
library(tidyverse)
realestate <- read.csv("real-estate.csv")
head(realestate)

```


```{r, comment=NA, message=FALSE, warning=FALSE}
glimpse(realestate)
```

\newpage

## Q1: Identify qualitative and quantitative variables.

```{r, comment=NA,message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=10}
library(GGally)
realestate$Airconditioning <- factor(realestate$Airconditioning)
realestate$Pool <- factor(realestate$Pool)
realestate$AdjHighway <- factor(realestate$AdjHighway)
realestate$Quality <- factor(realestate$Quality)
summary(realestate)
```

\newpage

## Q2: What is wrong with the following graph?

```{r, comment=NA,message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Pairwise correlation plot"}
realestate <- read.csv("real-estate.csv")
library(GGally)
ggpairs(realestate)

```

\newpage

## Q3: Figure 1 is modified as follows. Now what can you say about the graph?

```{r, comment=NA,message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Pairwise correlation plot"}
library(GGally)
realestate$Airconditioning <- factor(realestate$Airconditioning)
realestate$Pool <- factor(realestate$Pool)
realestate$AdjHighway <- factor(realestate$AdjHighway)
realestate$Quality <- factor(realestate$Quality)
ggpairs(realestate[, -c(1, 6, 8, 10, 12)])

```

\newpage

## Q4: What can you say about the source of multicollinearity in these data?

```{r, comment=NA, message=FALSE, warning=FALSE}
realestate$Airconditioning <- factor(realestate$Airconditioning)
realestate$Pool <- factor(realestate$Pool)
realestate$AdjHighway <- factor(realestate$AdjHighway)
realestate$Quality <- factor(realestate$Quality)
realestate <- realestate[, -1]
summary(realestate)
model <- lm(Price ~ . , data=realestate)
summary(model)
library(car)
car::vif(model)
```


\newpage

# Note: 

## 1.1: note

Multicollinearity occurrs if we do not treat this appropriately.


### 1.2 Model with only quantitative variables: VIF

```{r, comment=NA, message=FALSE}
summary(Duncan)
m1 <- lm(prestige ~ income + education, data=Duncan)
vif(m1)
```

### 1.3 Model with only quantitative variables and qualitative variables:  Generalized variance-inflation factors 

```{r, comment=NA, message=FALSE}
m2 <- lm(prestige ~ income + education + type, data=Duncan)
vif(m2)
```

\newpage


\newpage

## Q5: Write down the estimated model.

```{r, comment=NA, message=FALSE}
summary(model)
```

\newpage
