---
title: "STA 506 2.0 Linear Regression Analysis"
subtitle: "Lecture 9-i: Confidence Intervals in Multiple Regression"
author: "Dr Thiyanga S. Talagala"
date: "2020-10-31"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

## Recap


---
## Confidence Intervals in Multiple Regression

1. Confidence intervals on the regression coefficients

2. Confidence interval estimation of the mean response at a given point

---

## Confidence Intervals on the Regression Coefficients

To construct confidence intervals for regression coefficients ( $\beta_j$ , $j=0, 1, ...p$) we will continue to assume that,

- errors $\epsilon_i$ are normally and independently distributed with mean zero and variance $\sigma^2$. 

Hence, before constructing the confidence intervals you need to check the validity of the assumptions. 

---

## Confidence Intervals on the Regression Coefficients

Data set

```{r, comment=NA, message=FALSE, warning=FALSE}
library(tidyverse)
heart.data <- read_csv("heart.data.csv")
heart.data
```

---

## Confidence Intervals on the Regression Coefficients (cont.)

```{r, comment=NA, message=FALSE, warning=FALSE}

regHeart <-  lm(heart.disease ~ biking+ smoking, data=heart.data)
regHeart

```

Validity of the assumptions: All satisfied. We discussed in Week 8

### Compute 95% confidence intervals for regression coefficients


```{r, comment=NA, message=FALSE, warning=FALSE}
confint(regHeart, level=0.95)
```
---

### Compute 95% confidence intervals for regression coefficients


```{r, comment=NA, message=FALSE, warning=FALSE}
confint(regHeart, level=0.95)
```

### Compute 90% confidence intervals for regression coefficients


```{r, comment=NA, message=FALSE, warning=FALSE}
confint(regHeart, level=0.90)
```

---

### Interpretation of confidence intervals for regression coefficients


```{r, comment=NA, message=FALSE, warning=FALSE}
confint(regHeart, level=0.95)
```


---

### Confidence Interval Estimation of the Mean Response at a Given Point

$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \epsilon$$

where 

$Y$ - percentage of people with heart disease

$X_1$ - percentage of people in each town who bike to work

$X_2$ - percentage of people in each town who smoke

### Fitted regression model

```{r, comment=NA}
regHeart
```


$\hat{Y} = 14.9847 - 0.2001X_1 + 0.1783X_2$

where 

$\hat{Y}$ - Fitted values


---

### Confidence Interval Estimation of the Mean Response at a Given Point

Suppose we have an observation $X_1 = 60$ and $X_2=20$ and we would like to find a 95% confidence interval on the percentage of people with heart disease


**The fitted value at this point is:**

$\hat{Y} = 14.9847 - 0.2001X_1 + 0.1783X_2$

$\hat{Y} = 14.9847 - 0.2001(60) + 0.1783(20) = 6.54$


**A 95% confidence interval on the mean percentage of people with heart disease at this point is:**


```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='confidence', level=0.95)

```


**Interpretation: **

---

## Prediction of New Observation

Suppose we want to predict construct 95% prediction interval on the percentage people with heart disease at $X_1=60$ and $X_2=20$.

```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='predict', level=0.95)

```

<!--https://stats.stackexchange.com/questions/16493/difference-between-confidence-intervals-and-prediction-intervals-->

---
class: center, middle, duke-orange

# Confidence Interval Estimation of the 

# Mean Response vs. Prediction Interval 

Purpose: Illustrate the difference between confidence intervals for mean and prediction intervals.

---

### Prediction of the mean response

A 95% confidence interval on the **mean** percentage of people with heart disease at the point at $X_1=60$ and $X_2=20$.


```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='confidence', level=0.95)

```

### Prediction of a future value

Suppose we want to predict construct 95% **prediction interval** on the percentage people with heart disease at $X_1=60$ and $X_2=20$.

```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='predict', level=0.95)

```

---

.content-box-yellow[Prediction of the mean response]

what would be the average (mean) response with characteristics $X_1=60$ and $X_2=20$ ?

```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='confidence', level=0.95)

```

We predict the mean value of $Y$ with characteristics $X_1=60$ and $X_2=20$.

.content-box-yellow[Prediction of a future value]

what is the predicted value of $Y$ with characteristics $X_1=60$ and $X_2=20$?

```{r, comment=NA, message=FALSE, warning=FALSE}
predict(regHeart, list(biking = 60, smoking = 20),
interval='predict', level=0.95)

```

We predict $Y$ for a specific new case that comes from the population with characteristics $X_1=60$ and $X_2=20$.

---

## Prediction of Set of New Observations


```{r, comment=NA, message=FALSE}
newheartdata <- data.frame(biking = c(30, 40, 40, 60),
                           smoking = c(20, 30, 12, 10))
newheartdata
```


```{r, comment=NA, message=FALSE}
predict(regHeart, newdata=newheartdata , interval="predict")

```

---

class: center, middle



Acknowledgement

Introduction to Linear Regression Analysis, Douglas C. Montgomery, Elizabeth A. Peck, G. Geoffrey Vining


All rights reserved by 

[Dr. Thiyanga S. Talagala](https://thiyanga.netlify.app/) 



