---
title: "Simple Linear Regression Analysis with R"
author: "Prepared by Dr Thiyanga S. Talagala"
date: "26/09/2020"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, fig.height = 5, fig.width = 5)
```

# Load packages

```{r}
library(tidyverse)
library(broom)

```

# Load dataset

```{r, comment=NA}
house <- read_csv("house.csv")
attach(house)
```

## Overview of the dataset

```{r}
glimpse(house)
```


```{r}
summary(house)
```

# Visualise data

```{r}
qplot(data=house, x=price, geom=c("histogram"))+
  geom_histogram(color="black", fill="#d95f02")
```

Interpretation:


__________________


___________________

```{r}
qplot(data=house, x=houseSize, geom=c("histogram"))+
  geom_histogram(color="black", fill="#1b9e77")
```

Interpretation: 


___________________

___________________


```{r}
qplot(data=house, y=price, x=houseSize) + 
  xlab("House size (sqft)") + 
  ylab("Price") + 
  labs(title="Price vs House size (sqft)")
  
```


```{r}

cor(house$price, house$houseSize)
```

Interpretation: 


___________________

___________________





# Fit a regression model

```{r}
slrfit <- lm(price ~ houseSize, data=house)
slrfit

  
```

```{r}
qplot(data=house, y=price, x=houseSize) + 
  xlab("House size (sqft)") + 
  ylab("Price") + 
  labs(title="Price vs House size (sqft)") + 
   geom_abline(intercept = 5.15772, slope = 0.02489, colour="forestgreen", lwd=2) 
```

Write the fitted regression model:

___________________

___________________


# Model adequacy checking

## Compute fitted values and residuals

```{r}
slrfit_values <- augment(slrfit)
slrfit_values
```

## Residuals vs fitted values

Write the code to plot residual vs fitted values

```{r}
qplot(data=slrfit_values, y=.resid, x=.fitted)

```

___________________


## Normality test of residuals


```{r}
qplot(data=slrfit_values, x=.resid, geom=c("histogram"))+
  geom_histogram(color="black", fill="lightblue")
```

```{r}
ggplot(slrfit_values, 
       aes(sample=.resid))+
  stat_qq() + stat_qq_line()+labs(x="Theoretical Quantiles", y="Sample Quantiles")


```

```{r}
shapiro.test(house$price)
```

H0:

H1: 

Decision:

Conclusion: 




# Interpretation of the model output

```{r}
summary(slrfit)
```

## Coefficient of determination

$R^2 = 87.02\%$

Interpretation: 

___________________

___________________

___________________

___________________

## Point estimate of $\beta_0$

Interpretation: 


___________________

___________________


## Intervals estimate of $\beta_0$

Interpretation: 


___________________

___________________


## Point estimate of $\beta_1$

Interpretation: 


___________________

___________________


## Intervals estimate of $\beta_1$

Interpretation: 


___________________

___________________

# Hypothesis testing on the Slope and Intercept

___________________

___________________

___________________

___________________


___________________

___________________

___________________

___________________

___________________

___________________


___________________

___________________


___________________

___________________

___________________

___________________


___________________

___________________


___________________

___________________

___________________

___________________


___________________

___________________




# Prediction of New Observations


```{r, comment=NA}
newhouseSize <- data.frame(houseSize = c(2000, 2500, 2300, 2400, 2760))
predict(slrfit, newdata=newhouseSize , interval="predict")
```


