---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

## Exploratory Data Analysis - Summary Report



## 1. Data Overview

**Table 1: Composition of the sample**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
data(penguines)
summary(penguins)
```

According to the Table 1 the majority of penguins are Adelie and the majority of penguins are caught from Biscoe isaland. Except, in the variables species and island there are 2 missing values in other variables.


## 2. Composition of the sample

```{r, fig.cap="Composition of sample by Species", fig.height=2, fig.width=3}
ggplot(penguins) + geom_bar(aes(species))

```

\newpage

Sometimes, you will see row names as a variable. Then you will get multiple bars like this. Here is an example. See Figure 2. Do not include those types of charts to the report. 

```{r, echo=FALSE, fig.cap="Do not include these types ofgraphs "}
data("msleep")
ggplot(msleep) + geom_bar(aes(name))
```