---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

## Exploratory Data Analysis - Summary Report



## 1. Data Overview

**Table 1: Composition of the sample**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
data(penguines)
summary(penguins)
```

According to the Table 1 the majority of penguins are Adelie and the majority of penguins are caught from Biscoe isaland. Except, species and island there are 2 missing values in other variables.


## 2. Composition of the sample 



Sometimes table does not give the counts for the levels of the categorical variable. Then you can take bar charts to view individual counts. Here, is the command.

```{r, fig.cap="Composition of sample by Species", fig.height=2, fig.width=3}
ggplot(penguins) + geom_bar(aes(species))

```

You can do the same for other qualitative variables if necessary.

\newpage

**Important note:**

Sometimes, you will see row names as a variable. Then you will get multiple bars like this (Figure 2). Here is an example. See Figure 2. Do not include those types of charts to the report. 

```{r, echo=FALSE, fig.cap="Do not include these types ofgraphs", fig.height=4, fig.width=4}
data("msleep")
ggplot(msleep) + geom_bar(aes(name))
```

\newpage

**Composition of the sample by sex and island**

```{r, fig.cap="Composition of sample by species and gender"}
tablecount <- penguins %>%
  drop_na() %>%
  count(sex, species) 
tablecount
  ggplot(tablecount) + geom_col(aes(x = species, y = n, fill = species)) +
  geom_label(aes(x = species, y = n, label = n)) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  facet_wrap(~sex) +
  labs(title = 'Penguins Species ~ Gender')

```

According to Figure 3, the distribution of species type within male and female groups are approximately same.

## 3. Distribution of body characteristics variables by qualitative variables

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, fig.cap="Distribution of flipper length by gender"}
penguins2 <- penguins %>% drop_na()
ggplot(data = penguins2, 
       aes(y = flipper_length_mm,
           x = sex,
           fill=sex)) +
  geom_boxplot() 

```

According to figure 4 the flipper length of male penguins are higher than female penguins.

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, fig.cap="Distribution of flipper length by island"}
penguins2 <- penguins %>% drop_na()
ggplot(data = penguins2, 
       aes(y = flipper_length_mm,
           x = island,
           fill=island)) +
  geom_boxplot() 

```

According to figure 5 the flipper length of Biscoe is the highest when compared that with Dream and Torgersen penguins.

\newpage

## 4. Relationship between body-characteristics 



```{r, fig.cap="Relationship between body characteristics variables", warning=FALSE, message=FALSE}
body1 <- select(penguins, c(bill_length_mm, bill_depth_mm,
                           flipper_length_mm,
                           body_mass_g))
library(GGally)
ggpairs(body1)
```

According to the figure 6, there is a strong positive relationship between body mass and flipper length.

\newpage

```{r, fig.cap="Relationship between body characteristics variables by species", warning=FALSE, message=FALSE}
body2 <- select(penguins, c(bill_length_mm, bill_depth_mm,
                           flipper_length_mm,
                           body_mass_g, species))
ggpairs(data=body2, aes(color = species),
          columns = c("flipper_length_mm", "body_mass_g",
                      "bill_length_mm", "bill_depth_mm"))
```

According to the figure 7, the strength of the relationship between bill depth and flipper length vary according to the species type.

\newpage

```{r, fig.cap="Relationship between body characteristics variables by sex", warning=FALSE, message=FALSE}
body3 <- select(penguins, c(bill_length_mm, bill_depth_mm,
                           flipper_length_mm,
                           body_mass_g, sex))
ggpairs(data=body3, aes(color = sex),
          columns = c("flipper_length_mm", "body_mass_g",
                      "bill_length_mm", "bill_depth_mm"))
```

According to the Figure 8, there is a strong positive linear relationship between flipper length and body mass index. The pearson's correlation coefficients  between flipper length and body mass index are approximately same for female and male penguins.

\newpage

## Few other useful commands for plots.


```{r, comment=NA}
ggplot(penguins)+ 
  geom_point(aes(bill_length_mm, bill_depth_mm))+facet_wrap(~island)
```

```{r, comment=NA}
ggplot(penguins)+ 
  geom_point(aes(bill_length_mm, bill_depth_mm, col=species))
```
