---
title: Multicollinearity
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

Multicollinearity refers to a situation in which two or more **explanatory variables** in a multiple regression model are highly linearly related. 

## 2. Data

```{r, comment=NA, message=FALSE, warning=FALSE, message=FALSE}
library(readr)
bloodpressure <- read_csv("bloodpressure.csv")
bloodpressure
```

### 2.1 Variable description

1. $Y$: BP (blood pressure, in mmHg)

2. $X_1$: Age (in years)

3. $X_2$: Weight (in kg)

4. $X_3$: BSA (body surface area, in sq m)

5. $X_4$: Dur (duration of hypertension)

6.  $X_5$: Pulse (basal pulse)

7.  $X_6$: Stress (stress index)

## 3. How to detect multicollinearity?

1. Correlation matrix and scatterplot matrix

This is limiting. It is possible that the pairwise correlations between variables are small, but a linear dependence exists among three or even more variables in the dataset. Hence, we use **variance inflation factors (VIF)** to detect multicollinearity.

```{r, comment=NA, message=FALSE, warning=FALSE}
library(GGally)
ggpairs(bloodpressure[, -c(1, 2)])

```


2. Variance Inflation Factors (VIF)

Variance inflation factor for $j^{th}$ variable

$$VIF_j = \frac{1}{1-R^2_j}$$

where $R^2_j$ is the $R^2$ value obtained by regressing the $j^{th}$ predictor on the remaining predictors.


```{r, comment=NA, message=FALSE, warning=FALSE}
library(broom)
bp <- lm(BP ~ Age + Weight + BSA + Dur + Pulse + Stress, data=bloodpressure)
bp
summary(bp)
```

## 4. Calculate VIF

```{r, comment=NA, message=FALSE, warning=FALSE}
library(car)
vif(bp)
```

## 5. Illustration of the output for weight variable

Build a regression model taking *weight* as the dependent variable and remaining `x` variables as the independent variables.

```{r, comment=NA, message=FALSE, warning=FALSE}
weight <- lm(Weight ~ Age  + BSA + Dur + Pulse + Stress, data=bloodpressure)
weight
summary(weight)
```

$$VIF_weight = \frac{1}{1-R^2_weight} = \frac{1}{1-0.8812} = 8.42$$

**VIFs exceeding 4  indicates high multicollinearity while  VIFs exceeding 10 are considered  evidence of serious multicollinearity requiring correction.**

## 6. What to do now?

One solution is to remove some of the variables with high VIF. Variables `Weight`, `BSA` and `Pulse` have high VIF values.  If we review the pairwise correlations again, we can see `Weight` and `BSA` are highly correlated. We can choose to remove either predictor from the model. 

Which one to remove? In-class discussion.

```{r, comment=NA, message=FALSE, warning=FALSE}
library(GGally)
ggpairs(bloodpressure[, -c(1, 2)])

```

\newpage


## New model without `Pulse` and `BSA`

```{r, comment=NA, message=FALSE, warning=FALSE}
library(broom)
bp2 <- lm(BP ~ Age + Weight  + Dur  + Stress, data=bloodpressure)
bp2
summary(bp2)
vif(bp2)
```


## Ackowledgement

Data: The Pennsylvania State University.
